### 2018-03-14 submission/manuscript.Rmd
update title
introduction re-write

# High Fidelity DNA Polymerase can Introduce Bias into 16S rRNA Gene Sequencing Results

@@

It is important to differentiate between bias, reproducibility, and standardization since often time these three can be confused and used interchangeably with each other.

Need to talk about sources of bias (each of these could be a small paragraph)

  preservation kits (overgrowth, selective killing, DNA/RNA degradation at RT, freeze thaw)

  Within kits (e.g. contamination)

  PCR (16S hypervariable region primer specificity, amplification bias, error, polymerase dependent differences)

  Sequencing machines themselves (non-random indels, etc.)

Many of these sources of bias are common to both 16S rRNA gene and metagenomic sequencing approaches [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017]. However, here we focused primarily on cycle number and high fidelity (HiFi) DNA polymerase, two areas where introduced bias might have a larger influence on 16S rRNA gene sequencing than metagenomics.

*Talk about previous research on PCR bias*

A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) generated were not easily removed using the authors chosen bioinformatic pipeline  [@taq_Gohl_2016]. It is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

The most commonly studied biases are preservation media, storage conditions, and DNA extraction kits. Although DNA extraction kits have consistently been shown to add bias to downstream analysis the current literature on preservation media and storage conditions has been mixed, with some studies showing biases while others do not [@contamination_Salter2014; @metagenomcis_bias_Costea_2017; @preservation_Song_2016; @preservation_Luo_2016; @preservation_Dominianni_2014; @storage_Lauber_2010; @storage_Bassis_2017; @preservation_Sinha_2015]. The study of these specific biases has become so large, aggregating them all together has become a difficult task with some researchers providing resources to actively track new findings (e.g. Microbiome Digest - https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/). It is evident that these three sources of bias have been extensively covered within the literature but they are not the only sources that should be critically examined. The type of DNA polymerase chosen could also have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.

>>>>

# High Fidelity DNA Polymerase Introduces Bias into 16S rRNA Gene Sequencing Results

@@

It is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other. Bias can change the observed results in a way that is reproducible and standardized. For example, DNA extraction kits used for 16S rRNA gene sequencing biases your findings versus another group not using the same brand kit but both groups have reproducible results. Therefore, standardization of 16S rRNA gene sequencing methods to increase reproducibility can still be problematic due to bias. Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies within the context of the overall field.

A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [ref]. Similarily, this overgrowth could still occur if the preservation media does not adequately inhibit growth [ref]. Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [ref]. Additionally, reagent contamination has been shown to add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014]. Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [refs]. However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method used.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that primers used, cycle number, selective amplification, amplification error, non-specific amplification, chimeras, and polymerase dependent differences can all change the observed results [refs]. Many of these sources of biases are made worse as cycle number increases. The selective amplification of the dominant sequence can increase the difference between the most and next abundant sequence, over-emphasizing the actual difference [ref]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) also increases as cycle number increases [refs]. Further, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension [@Haas2011]. Even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen. There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias. With respect to primers, there are different binding affinities for different bacteria that are variable region dependent [refs]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014]. Interestingly, the influence these different DNA polymerases can have on the observed results have not been well studied like the previously mentioned sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline  [@taq_Gohl_2016]. It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on HiFi DNA polymerases. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

### 2018-03-14 submission/manuscript.tex
introduction re-write

\section{The Influence of High Fidelity DNA Polymerase on 16S rRNA Gene
Sequencing}\label{the-influence-of-high-fidelity-dna-polymerase-on-16s-rrna-gene-sequencing}

@@

\textbf{Background.} Research has identified various methodological
steps along the 16S rRNA gene survey pipeline that can change microbial
community results. Although cycle number and high fidelity (HiFi) DNA
polymerase are studied less often, they are still important sources of
bias. Here, we critically examine both cycle number and HiFi DNA
polymerase for biases that may influence downstream bacterial community
results.

@@

Recently, there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). Investigating ways that 16S rRNA gene
surveys can be made more reproducible has been a predominant focal point
of this push towards standardization (Lauber et al., 2010; Salter et
al., 2014; Song et al., 2016; Gohl et al., 2016). Due to this focus, 16S
rRNA gene sequencing has been much maligned for introduced biases. Yet,
many of these same considerations also affect metagenomic sequencing
(Nayfach \& Pollard, 2016; Costea et al., 2017). Between the two
approaches similar bias considerations include preservation media,
storage conditions, DNA extraction kit, PCR, and sequence library
preparation. For these overlapping considerations, biases identified for
16S rRNA gene sequencing will also likely influence metagenomic
sequencing results.

The most commonly studied biases are preservation media, storage
conditions, and DNA extraction kits. Although DNA extraction kits have
consistently been shown to add bias to downstream analysis the current
literature on preservation media and storage conditions has been mixed,
with some studies showing biases while others do not (Lauber et al.,
2010; Dominianni et al., 2014; Salter et al., 2014; Sinha et al., 2015;
Song et al., 2016; Luo et al., 2016; Bassis et al., 2017; Costea et al.,
2017). The study of these specific biases has become so large,
aggregating them all together has become a difficult task with some
researchers providing resources to actively track new findings
(e.g.~Microbiome Digest -
\url{https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/}).
It is evident that these three sources of bias have been extensively
covered within the literature but they are not the only sources that
should be critically examined. The type of DNA polymerase chosen could
also have a wide ranging affect on downstream results due to error rates
and chimeras that may not be easily resolved using bioinformatic
approaches.

A recent study in \emph{Nature Biotechnology} found clear differences
between normal and high fidelity (HiFi) DNA polymerase and that you
could reduce error and chimera generation by optimizing the PCR protocol
(Gohl et al., 2016). This study also found that, regardless of DNA
polymerase, the number of Operational Taxonomic Units (OTUs) generated
were not easily removed using the authors chosen bioinformatic pipeline
(Gohl et al., 2016). It is natural to extend this line of inquiry and
ask whether different HiFi DNA polymerase contribute different biases to
downstream sequencing results. There is some reason to think that this
may be the case since many of these HiFi DNA polymerase come from
different families (e.g. \emph{Taq} belongs to the family A polymerases)
and may intrinsically have different error rates that cannot be
completely removed with modifications (Ishino \& Ishino, 2014). In this
study we critically examine if any of five different HiFi DNA
polymerases introduce significant biases into 16S rRNA gene surveys, if
this is a cycle dependent phenomenon, and whether they can be removed
using a standard bioinformatic pipeline.

@@

\hypertarget{refs}{}
\hypertarget{ref-storage_Bassis_2017}{}
Bassis CM., Nicholas M. Moore., Lolans K., Seekatz AM., Weinstein RA.,
Young VB., Hayden MK. 2017. Comparison of stool versus rectal swab
samples and storage conditions on bacterial community profiles.
\emph{BMC Microbiology} 17. DOI:
\href{https://doi.org/10.1186/s12866-017-0983-9}{10.1186/s12866-017-0983-9}.

\hypertarget{ref-benjamini_controlling_1995}{}
Benjamini Y., Hochberg Y. 1995. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. \emph{Journal of
the Royal Statistical Society. Series B (Methodological)} 57:289--300.

@@

\hypertarget{ref-metagenomcis_bias_Costea_2017}{}
Costea PI., Zeller G., Sunagawa S., Pelletier E., Alberti A., Levenez
F., Tramontano M., Driessen M., Hercog R., Jung F-E., Kultima JR.,
Hayward MR., Coelho LP., Allen-Vercoe E., Bertrand L., Blaut M., Brown
JRM., Carton T., Cools-Portier S., Daigneault M., Derrien M., Druesne
A., Vos WM de., Finlay BB., Flint HJ., Guarner F., Hattori M., Heilig
H., Luna RA., Hylckama Vlieg J van., Junick J., Klymiuk I., Langella P.,
Chatelier EL., Mai V., Manichanh C., Martin JC., Mery C., Morita H.,
O'Toole PW., Orvain C., Patil KR., Penders J., Persson S., Pons N.,
Popova M., Salonen A., Saulnier D., Scott KP., Singh B., Slezak K.,
Veiga P., Versalovic J., Zhao L., Zoetendal EG., Ehrlich SD., Dore J.,
Bork P. 2017. Towards standards for human fecal sample processing in
metagenomic studies. \emph{Nature Biotechnology}. DOI:
\href{https://doi.org/10.1038/nbt.3960}{10.1038/nbt.3960}.

\hypertarget{ref-preservation_Dominianni_2014}{}
Dominianni C., Wu J., Hayes RB., Ahn J. 2014. Comparison of methods for
fecal microbiome biospecimen collection. \emph{BMC Microbiology} 14:103.
DOI:
\href{https://doi.org/10.1186/1471-2180-14-103}{10.1186/1471-2180-14-103}.

@@

\hypertarget{ref-review_Hugerth_2017}{}
Hugerth LW., Andersson AF. 2017. Analysing microbial community
composition through amplicon sequencing: From sampling to hypothesis
testing. \emph{Frontiers in Microbiology} 8. DOI:
\href{https://doi.org/10.3389/fmicb.2017.01561}{10.3389/fmicb.2017.01561}.

@@

\hypertarget{ref-review_Kim_2017}{}
Kim D., Hofstaedter CE., Zhao C., Mattei L., Tanes C., Clarke E., Lauder
A., Sherrill-Mix S., Chehoud C., Kelsen J., Conrad M., Collman RG.,
Baldassano R., Bushman FD., Bittinger K. 2017. Optimizing methods and
dodging pitfalls in microbiome research. \emph{Microbiome} 5. DOI:
\href{https://doi.org/10.1186/s40168-017-0267-5}{10.1186/s40168-017-0267-5}.

@@

\hypertarget{ref-storage_Lauber_2010}{}
Lauber CL., Zhou N., Gordon JI., Knight R., Fierer N. 2010. Effect of
storage conditions on the assessment of bacterial community structure in
soil and human-associated samples. \emph{FEMS Microbiology Letters}
307:80--86. DOI:
\href{https://doi.org/10.1111/j.1574-6968.2010.01965.x}{10.1111/j.1574-6968.2010.01965.x}.

\hypertarget{ref-preservation_Luo_2016}{}
Luo T., Srinivasan U., Ramadugu K., Shedden KA., Neiswanger K., Trumble
E., Li JJ., McNeil DW., Crout RJ., Weyant RJ., Marazita ML., Foxman B.
2016. Effects of specimen collection methodologies and storage
conditions on the short-term stability of oral microbiome taxonomy.
\emph{Applied and Environmental Microbiology} 82:5519--5529. DOI:
\href{https://doi.org/10.1128/aem.01132-16}{10.1128/aem.01132-16}.

@@

\hypertarget{ref-metagenomcis_bias_Nayfach_2016}{}
Nayfach S., Pollard KS. 2016. Toward accurate and quantitative
comparative metagenomics. \emph{Cell} 166:1103--1116. DOI:
\href{https://doi.org/10.1016/j.cell.2016.08.007}{10.1016/j.cell.2016.08.007}.

@@

\hypertarget{ref-preservation_Sinha_2015}{}
Sinha R., Chen J., Amir A., Vogtmann E., Shi J., Inman KS., Flores R.,
Sampson J., Knight R., Chia N. 2015. Collecting fecal samples for
microbiome analyses in epidemiology studies. \emph{Cancer Epidemiology
Biomarkers \& Prevention} 25:407--416. DOI:
\href{https://doi.org/10.1158/1055-9965.epi-15-0951}{10.1158/1055-9965.epi-15-0951}.

\hypertarget{ref-preservation_Song_2016}{}
Song SJ., Amir A., Metcalf JL., Amato KR., Xu ZZ., Humphrey G., Knight
R. 2016. Preservation methods differ in fecal microbiome stability,
affecting suitability for field studies. \emph{mSystems} 1:e00021--16.
DOI:
\href{https://doi.org/10.1128/msystems.00021-16}{10.1128/msystems.00021-16}.

>>>>

\section{High Fidelity DNA Polymerase Introduces Bias into 16S rRNA Gene
Sequencing
Results}\label{high-fidelity-dna-polymerase-introduces-bias-into-16s-rrna-gene-sequencing-results}

@@

\textbf{Background.} Many of the different steps used during 16S rRNA
gene sequencing can introduce bias and by extension change the observed
microbial community results. Although cycle number and high fidelity
(HiFi) DNA polymerase are studied less often, they are still important
sources of bias in this workflow. Here, we critically examine both cycle
number and HiFi DNA polymerase for biases that may influence downstream
bacterial community results.

@@

It is important to differentiate between bias, reproducibility, and
standardization since often times these three can be confused and used
interchangeably with each other. Bias can change the observed results in
a way that is reproducible and standardized. For example, DNA extraction
kits used for 16S rRNA gene sequencing biases your findings versus
another group not using the same brand kit but both groups have
reproducible results. Therefore, standardization of 16S rRNA gene
sequencing methods to increase reproducibility can still be problematic
due to bias. Determining how different aspects of a 16S rRNA gene
sequencing workflow could bias the observed results is critical for the
interpretation of specific studies within the context of the overall
field.

A typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. Generally, not
using a preservation media and leaving samples at room temperature has
been shown to cause overgrowth of low abundance members of the fecal
bacterial community {[}ref{]}. Similarily, this overgrowth could still
occur if the preservation media does not adequately inhibit growth
{[}ref{]}. Reports have also shown that changes in specific community
members might occur due to differing susceptibility to freeze thaw
cycles amongst microbes {[}ref{]}. Additionally, reagent contamination
has been shown to add community members and the contribution of these
contaminant members grows larger with lower biomass samples (Salter et
al., 2014). Recent studies have shown that the majority of these biases
due to either preservation or extraction tend to be smaller than the
overall biological signal being measured {[}refs{]}. However, the
contribution of PCR bias to this overall workflow is not well
characterized since these studies use the same PCR approach while
varying preservation or extraction method used.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that primers used,
cycle number, selective amplification, amplification error, non-specific
amplification, chimeras, and polymerase dependent differences can all
change the observed results {[}refs{]}. Many of these sources of biases
are made worse as cycle number increases. The selective amplification of
the dominant sequence can increase the difference between the most and
next abundant sequence, over-emphasizing the actual difference
{[}ref{]}. Both amplification error and non-specific amplification
(e.g.~incorrect amplicon size products) also increases as cycle number
increases {[}refs{]}. Further, chimeras can form from an aborted
extension followed by subsequent priming error and secondary extension
(Haas et al., 2011). Even if it has not been explicitly shown to
increase with cycle number it is easy to postulate that this does
happen. There are also intrinsic properties to primers and DNA
polymerases chosen that can introduce bias. With respect to primers,
there are different binding affinities for different bacteria that are
variable region dependent {[}refs{]}. Additionally, there are multiple
families of DNA polymerases that have their own error rate and proof
reading capcity (Ishino \& Ishino, 2014). Interestingly, the influence
these different DNA polymerases can have on the observed results have
not been well studied like the previously mentioned sources of PCR-based
bias.

A recent study found clear differences between normal and high fidelity
(HiFi) DNA polymerase and that you could reduce error and chimera
generation by optimizing the PCR protocol (Gohl et al., 2016). This
study also found that, regardless of DNA polymerase, the number of
Operational Taxonomic Units (OTUs) or taxa generated were not easily
reduced using the authors chosen bioinformatic pipeline (Gohl et al.,
2016). It is natural to extend this line of inquiry and ask if biases in
the number of OTUs and chimeras are also dependent on HiFi DNA
polymerases. There is some reason to think that this may be the case
since many of these HiFi DNA polymerase come from different families
(e.g. \emph{Taq} belongs to the family A polymerases) and may
intrinsically have different error rates that cannot be completely
removed with modifications (Ishino \& Ishino, 2014). In this study we
critically examine if any of five different HiFi DNA polymerases
introduce significant biases into 16S rRNA gene surveys, if this is a
cycle dependent phenomenon, and whether they can be removed using a
standard bioinformatic pipeline.

@@

\hypertarget{refs}{}
\hypertarget{ref-benjamini_controlling_1995}{}
Benjamini Y., Hochberg Y. 1995. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. \emph{Journal of
the Royal Statistical Society. Series B (Methodological)} 57:289--300.

@@

@@

\hypertarget{ref-Haas2011}{}
Haas BJ., Gevers D., Earl AM., Feldgarden M., Ward DV., Giannoukos G.,
Ciulla D., Tabbaa D., Highlander SK., Sodergren E., Methe B., DeSantis
TZ., Petrosino JF., Knight R., and BWB. 2011. Chimeric 16S rRNA sequence
formation and detection in sanger and 454-pyrosequenced PCR amplicons.
\emph{Genome Research} 21:494--504. DOI:
\href{https://doi.org/10.1101/gr.112730.110}{10.1101/gr.112730.110}.

@@

@@

@@

@@
